---
title: 'How Does New Competition Affect Sales?'
author: "ggep"
date: '2015-12-13'
output:
  html_document:
    number_sections: yes
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
---


***
# Motivation

This script is intended to explore how competition affects
Rossmann store sales --- and in particular, what the effect
of new competition looks like.

Findings here may be useful in building predictive models for
the accompanying [Kaggle's Rossmann sales forecasting competition](https://www.kaggle.com/c/rossmann-store-sales).
They may also hold helpful business insights for Rossmann
in deciding new store sales location and anticipating reactions
when new competition opens.
As an [RMarkdown](http://rmarkdown.rstudio.com/) file, all steps
here are reproducible through Kaggle's Rossman [data](https://www.kaggle.com/c/rossmann-store-sales/data)
and the code below.

This script also highlights functionality of different R packages.
Packages used here:

```{r echo=FALSE}
rm(list=ls())
```
```{r results='hide', message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(scales)
```


***
# Data Preparation

This data is unique at both a date-level (990 days from 1/1/13
until 9/17/15) and store-level (1,115 total stores).  The analysis
to come focuses first at the store-level (where the competition data
are) but also moves to the date-level for understanding effects
of new competition on sales over time.

First, load train, test, and store data (`dplyr::read_csv` is faster than `utils::read.table`):

```{r results='hide', message=FALSE}
tr <- read_csv("../input/train.csv", col_types = list(StateHoliday = col_character()))
te <- read_csv("../input/test.csv", col_types = list(StateHoliday = col_character()))
st <- read_csv("../input/store.csv")
```

Add each store's average sales and customers to the store-level
data frame:

```{r}
st <- left_join(st, summarize(group_by(tr[tr$Open == 1,], Store)
                              , AvgSales = mean(Sales, na.rm = TRUE)
                              , AvgCust = mean(Customers, na.rm = TRUE))
                , "Store")
```

Add additional fields and clean-up:

```{r warning=FALSE}
st$Assortment[st$Assortment == "a"] <- "basic"
st$Assortment[st$Assortment == "b"] <- "extra"
st$Assortment[st$Assortment == "c"] <- "extended"
st$StTyAs <- paste("'", st$StoreType, "'", " store type - "
                   , st$Assortment, sep = "")
st$CompDistBin <- ntile(st$CompetitionDistance, 5)
st$CompDistBin <- paste(st$CompDistBin, "%ile")
st$CompDistBin[st$CompDistBin == "NA %ile"] <- NA
st$CompOpenDate <- ymd(paste(st$CompetitionOpenSinceYear
                             , st$CompetitionOpenSinceMonth
                             , "01"))
st$CompOpenDate <- as.Date(st$CompOpenDate)
st <- select(st
             , Store
             , StoreType
             , Assortment
             , StTyAs
             , CompDist = CompetitionDistance
             , CompDistBin
             , CompOpenDate
             , AvgSales
             , AvgCust)
st$AvgSpC <- st$AvgSales / st$AvgCust
st$CompDays <- as.numeric(as.Date("2015-09-17") - st$CompOpenDate)
```

One note to keep in mind for upcoming analysis:
while a competition distance is provided for all but three
of the 1,115 stores,
the competition open date variable is missing for 354 stores.


***
# Competition Effects

One place to start is the distributions of competition openings 
and competition distance --- the two variables provided to account
for competition effects.

Competition openings:

```{r warning=FALSE}
ggplot(st[st$StoreType != "b",]
       , aes(CompOpenDate, fill = StoreType)) + 
  stat_bin(binwidth = 365) +
  scale_x_date("Date of Competition Opening") +
  scale_y_continuous("# of Stores")
```

There was one competition store that opened January 1900 (store 815) 
and another that opened October 1961 (store 146), while the majority of others
opened after 2000. This may reflect reality (Rossmann either had
very few of their own stores or very few competition for a long
period of time) or data quality.  Store type doesn't appear to
necessarily have different competition openings.

Competition distance:

```{r warning=FALSE}
lbrk <- c(100, 1000, 10000)
ggplot(st[st$StoreType != "b",]
       , aes(CompDist, fill = StoreType)) + 
  geom_density(adjust = .7, alpha = .5) +
  scale_x_log10("Competition Distance", breaks = lbrk, label = comma) +
  scale_y_continuous("# of Stores")
```

Note the log axis; competition can be distance is skewed-right
and this transformation for distance may be appropriate.
The distribution appears fairly consistent among store types,
although "d" stores may have competition slightly further away.
Also, there did not appear to be any noticeable interactions between
competition opening and competition distance.

How does competition distance affect customers and average
sales per customer?

```{r warning=FALSE}
int_TilePlot <- function(data, response, x, y
                         , r_lab = response, x_lab = x, y_lab = y
                         , r_scale = comma, x_scale = comma, y_scale = comma
                         , bins = 10, exp = 8){
  library(ggplot2)
  library(scales)
  x_brk <- seq(min(data[[x]]), max(data[[x]]), length.out = bins)
  y_brk <- seq(min(data[[y]]), max(data[[y]]), length.out = bins)
  df <- data.frame(x_val = rep(x_brk, bins)
                   , y_val = rep(y_brk, each = bins)
                   , r_val = NA)
  df$x_z <- (df$x_val - mean(data[[x]])) / sd(data[[x]])
  df$y_z <- (df$y_val - mean(data[[y]])) / sd(data[[y]])
  data$x_z <- (data[[x]] - mean(data[[x]])) / sd(data[[x]])
  data$y_z <- (data[[y]] - mean(data[[y]])) / sd(data[[y]])
  for (i in 1:nrow(df)){
    dist <- numeric()
    for (j in 1:nrow(data)){
      dist <- append(dist, sqrt(abs(df$x_z[i] - data$x_z[j])^2
                                + abs(df$y_z[i] - data$y_z[j])^2))
    }
    adjDist <- abs(dist - max(dist))^exp
    wtdResp <- sum(adjDist * data[[response]]) / sum(adjDist)
    df$r_val[i] <- wtdResp
  }
  ggplot(df, aes(x = x_val, y = y_val, z = r_val)) +
    geom_tile(aes(fill = r_val)) +
    stat_contour(bins = 15) +
    scale_x_continuous(x_lab, labels = x_scale) +
    scale_y_continuous(y_lab, labels = y_scale) +
    scale_fill_continuous(r_lab, labels = r_scale) +
    theme_minimal()
}
int_TilePlot(st[st$StoreType != 'b' & !is.na(st$CompDist),]
             , "CompDist", "AvgSpC", "AvgCust"
             , "Competition Distance"
             , "Average Sales per Customer"
             , "Average # of Customers"
             , comma, dollar, comma)
```

The heatmap above shows stores with competition close by
are in the upper-left; they average a larger number of customers
but those customers have a fewer number of average sales.
This makes sense that stores with competition further away
are likely more remote and thus frequented by a smaller number
of customers, but when those customers do come, they average
a higher number of sales per customers.

The following section explores the effect of new competition
(competition date) on sales.

# New Competition

In order to understand the effect of new competition on sales
over time, we need to first do some more data preparation,
moving from the store-level data back into the larger
date- and store-level data.

```{r}
# join store new variables back on training data
tr <- left_join(tr, select(st, -AvgSales, -AvgCust, -AvgSpC)
                , "Store")
# create competition comparison data frame when sales data is
# available for new stores
cc <- tr[!is.na(tr$CompOpenDate)
         & tr$CompOpenDate > as.Date("2013-01-01")
         & tr$CompOpenDate < as.Date("2015-07-31")
         & !is.na(tr$Sales)
         & tr$Sales != 0
         & tr$StoreType != "b"
         & tr$StTyAs != "'c' store type  - extended",]
# calculate additional fields
cc$AvgSpC <- cc$Sales / cc$Customers
cc$IsCompOpen <- ifelse(cc$Date >= as.Date(cc$CompOpenDate), 1, 0)
cc$IsCompOpen[!is.na(cc$CompDist) & is.na(cc$CompOpenDate)] <- 1
cc$IsCompOpen[is.na(cc$CompDist)] <- NA
cc$DaysSinceNewComp <- as.numeric(cc$Date - cc$CompOpenDate)
cc <- cc[abs(cc$DaysSinceNewComp) < 365,]
# plot new store openings
ggplot(cc, aes(CompOpenDate)) +
  stat_bin(binwidth = 7) +
  scale_x_date("Competition Open Date") +
  scale_y_continuous("# of New Competition Day-Store Observations"
                     , labels = comma)
```

The above plot shows that new competition for Rossmann appears to
have arrived fairly randomly for the period in which we have sales
data.

Now we can see if there is a noticeable trend after stores open:

```{r message=FALSE}
ggplot(cc, aes(DaysSinceNewComp, Customers
               , col = as.factor(IsCompOpen))) +
  geom_point(alpha = .2) +
  stat_smooth(se = FALSE, span = .5, size = 1.5, col = "gray50") +
  scale_x_continuous("# of Days After (Before) New Competition") +
  scale_y_continuous("# of Daily Store Customers", labels = comma) +
  scale_color_discrete("Is Comp. Open?")
```

It appears there may be a small uptick in sales before new competition
opens and a downturn after it opens, but there is a lot of variation
within the data to notice a strong trend.

As such, building very simple model sales prediction models --- which do
not include any competition effects --- and looking at the residuals can help
isolate the effects of competition and bring all stores to a more
similar "level".

```{r message=FALSE}
# build simple linear regression models
m_cust <- lm(Customers ~ Promo + as.factor(DayOfWeek) + StateHoliday
             + as.factor(Store), cc)
m_aspc <- lm(AvgSpC ~ Promo + as.factor(DayOfWeek) + StateHoliday
             + as.factor(Store), cc)
# compute residuals
cc$res_m_cust <- cc$Customers - predict(m_cust)
cc$res_m_aspc <- cc$AvgSpC - predict(m_aspc)
# aggregate by median residuals for smoother patterns
cca <- summarize(group_by(cc, DaysSinceNewComp)
                 , NumObs = length(Date)
                 , IsCompOpen = max(IsCompOpen)
                 , MedResCust = median(res_m_cust)
                 , MedResASpC = median(res_m_aspc))
# new plot
ggplot(cca, aes(DaysSinceNewComp, MedResCust
               , col = as.factor(IsCompOpen))) +
  stat_smooth(se = FALSE, size = 1.5) +
  scale_x_continuous("# of Days After (Before) New Competition") +
  scale_y_continuous("Residual of # of Daily Store Customers"
                     , labels = comma) +
  scale_color_discrete("Is Comp. Open?") +
  geom_hline(y = 0, lty = 2, col = "gray20") +
  geom_vline(x = 0, lty = 2, col = "gray20")
```

The plot above now shows a possible trend of more customers
buying at Rossmann before competition opens, then an increasing
drop for the first 100 days once it opens.  The number of
customers start to level off, before dropping again around 225
days since new competition --- this last drop-off may be an artifact
of smaller sample size, or could be a true effect.  Further analysis
(such as with more historical data)
may be appropriate in fully understanding the longer-term effects of
new competition.

How does does competition distance affect the opening of a new store?

```{r message=FALSE}
# aggregate by median residuals for smoother patterns
ccac <- summarize(group_by(cc, DaysSinceNewComp, CompDistBin)
                 , NumObs = length(Date)
                 , IsCompOpen = max(IsCompOpen)
                 , MedResCust = median(res_m_cust)
                 , MedResASpC = median(res_m_aspc))
# new plot
ggplot(ccac, aes(DaysSinceNewComp, MedResCust
               , col = as.factor(CompDistBin))) +
  stat_smooth(se = FALSE, span = .6, size = 1.5) +
  scale_x_continuous("# of Days After (Before) New Competition") +
  scale_y_continuous("Residual of # of Daily Store Customers"
                     , labels = comma) +
  scale_color_discrete("Comp. Dist. %ile") +
  geom_hline(y = 0, lty = 2, col = "gray20") +
  geom_vline(x = 0, lty = 2, col = "gray20")
```

The pattern here stands out.  Stores that have the closest competition
are in the 1st percentile, and the effect of new competition opening up
for them is by far the strongest.  The effects decrease in magnitude
for the 2nd and 3rd percentiles, respectively.  In contrast, the 4th
and 5th percentiles --- with competition the furthest away --- appear
to not be affected at all when new competition opens.

Overall, new competition and that competition distance strongly
affect Rossmann sales. This insight can be valuable for both 
building sales forecasting models as business strategy in
planning store locations, anticipating new competition, and
reacting to new competition.

