---
title: "test"
author: "Bikash Agrawal"
date: "30 Oct 2015"
output: html_document
---

This is an exploratory analysis of the Rossmann Store Sales data which can be 
found [here](https://www.kaggle.com/c/rossmann-store-sales). The data isn't huge
but the speedup using data.table is noticeable. It is nice to have unmasked data which allows for some interpretations.
rmarkdown::render("test.Rmd")

Read in the data:

```{r}
library(data.table)
library(zoo)
library(forecast)
library(ggplot2)
train <- read.table("data/train.csv",sep=',',header = T)
test <- read.table("data/test.csv",sep=',',header = T)
store <- read.table("data/store.csv",sep=',',header = T)
```

Let's have a first look at the data:

```{r}
str(train)
str(test)
str(store)
# head(train); tail(train)
# head(test); tail(test)
#train[, Date := as.Date(Date)]
#test[, Date := as.Date(Date)]
train$Date <- as.Date(train$Date)
test$Date <- as.Date(test$Date)
store
#train <- train[order(Date)]
#test <- test[order(Date)]
summary(train)
summary(test)
test[is.na(test$Open), ] # Only store 622
test$Open[test$Store == 622]
```

The test set has just 41088 rows while the train set has 1017209 rows. 
The public leaderboard is based on 39% of the data (16024 rows) and the private leaderboard
is based on 61% of the data (25064 rows). Store 622 has 11 missing values in the
`Open` columns, but not all of the data in that column of that store is missing.
As was pointed out [here](https://www.kaggle.com/c/rossmann-store-sales/forums/t/16835/open-is-blank-in-test-file-for-store-622) it should probably be imputed as 1.

Additionally, the whole `Customers` column is missing from the test data (since
that data is only known ex post).

```{r}
test[is.na(test)] <- 1
```

During the test period there are no Easter or Christmas holidays but interestingly
during a rather large portion of the time (44%) there are school holidays while that
is the case for only 18% of the train data:

```{r}
# Unique values per column
train[, lapply(.SD, function(x) length(unique(x)))]
test[, lapply(.SD, function(x) length(unique(x)))]
# All test stores are also in the train data
sum(unique(test$Store) %in% unique(train$Store)) 
# 259 train stores are not in the test data
sum(!(unique(train$Store) %in% unique(test$Store))) 
table(train$Open) / nrow(train) # Percent Open Train
table(test$Open) / nrow(test) # Percent Open Test 
table(train$Promo) / nrow(train) # Percent of the time promo in train
table(test$Promo) / nrow(test) # Percent of the time promo in test
table(train$StateHoliday) / nrow(train) # Percent of the time holiday in train
table(test$StateHoliday) / nrow(test) # no b and c = no easter holiday and no christmas
table(train$SchoolHoliday) / nrow(train) # Percent of the time school holiday in train
table(test$SchoolHoliday) / nrow(test) # Percent of the time school holiday in test
```

There are no obvious breaks in the data.
The test period ranges from 2015-08-01 to 2015-09-17, so the task is to predict 48 days.
The train period ranges from 2013-01-01 to 2015-07-31.

```{r}
plot(train$Date, type = "l")
plot(test$Date, type = "l")
# As expected all 856 stores to be predicted daily
all(table(test$Date) == 856) 
```

Let's look at the columns that are unique to the train set:

```{r}
hist(train$Sales, 100)
hist(aggregate(train[Sales != 0]$Sales, 
               by = list(train[Sales != 0]$Store), mean)$x, 100, 
     main = "Mean sales per store when store was not closed")

hist(train$Customers, 100)
hist(aggregate(train[Sales != 0]$Customers, 
               by = list(train[Sales != 0]$Store), mean)$x, 100,
     main = "Mean customers per store when store was not closed")
ggplot(train[Sales != 0], aes(x = factor(SchoolHoliday), y = Sales)) +
    geom_jitter(alpha = 0.1) +
    geom_boxplot(color = "yellow", outlier.colour = NA, fill = NA)
ggplot(train[train$Sales != 0 & train$Customers != 0],
       aes(x = log(Customers), y = log(Sales))) + 
    geom_point(alpha = 0.2) + geom_smooth()
ggplot(train[train$Sales != 0 & train$Customers != 0],
       aes(x = factor(Promo), y = Sales)) + 
    geom_jitter(alpha = 0.1) +
    geom_boxplot(color = "yellow", outlier.colour = NA, fill = NA)
ggplot(train[train$Sales != 0 & train$Customers != 0],
       aes(x = factor(Promo), y = Customers)) + 
    geom_jitter(alpha = 0.1) +
    geom_boxplot(color = "yellow", outlier.colour = NA, fill = NA)
```

Note: I chose to not plot that data including days with 0 sales 
or customers because that would have biased the boxplots.

Sales is as expected strongly correlated with the number of customers.
It looks like the Boxplots of customers overlap a little more than the boxplots
of sales. This would mean that the promos are not mainly attracting more customers
but make customers spend more. The mean amount spent per customer is about one
Euro higher:

```{r}
with(train[train$Sales != 0 & train$Promo == 0], mean(Sales / Customers))
with(train[train$Sales != 0 & train$Promo == 1], mean(Sales / Customers))
```

There are sometimes promos while the respective store is closed and there are promos
45% of the time:

```{r}
table(ifelse(train$Sales != 0, "Sales > 0", "Sales = 0"),
      ifelse(train$Promo, "Promo", "No promo"))
```

At least there are no sales when the stores are closed but there are some stores
that, according to the data, made no sales although they were opened even if 
they had some customers. These observations *may* be errors in the data / outliers:

```{r}
table(ifelse(train$Open == 1, "Opened", "Closed"),
      ifelse(train$Sales > 0, "Sales > 0", "Sales = 0"))
# That tends to happen on consecutive days. Some stores even had customers
# (who bought nothing?)
train[Open == 1 & Sales == 0]
```

The stores have different amounts of days with zero sales.
There are spikes in the sales before the stores close and after the reopen:

```{r}
zerosPerStore <- sort(tapply(train$Sales, list(train$Store), function(x) sum(x == 0)))
hist(zerosPerStore,100)
# Stores with the most zeros in their sales:
tail(zerosPerStore, 10)
# Some stores were closed for some time, some of those were closed multiple times
plot(train[Store == 972, Sales], ylab = "Sales", xlab = "Days", main = "Store 972")
plot(train[Store == 103, Sales], ylab = "Sales", xlab = "Days", main = "Store 103")
plot(train[Store == 708, Sales], ylab = "Sales", xlab = "Days", main = "Store 708")
```

There are also stores that have *no* zeros in their sales. These are the exception
since they are opened also on sundays / holidays. The sales of those stores
on sundays are particularly high:

```{r}
ggplot(train[Store == 85], 
       aes(x = Date, y = Sales, 
           color = factor(DayOfWeek == 7), shape = factor(DayOfWeek == 7))) + 
    geom_point(size = 3) + ggtitle("Sales of store 85 (True if sunday)")
ggplot(train[Store == 262], 
       aes(x = Date, y = Sales, 
           color = factor(DayOfWeek == 7), shape = factor(DayOfWeek == 7))) + 
    geom_point(size = 3) + ggtitle("Sales of store 262 (True if sunday)")
```

That is not true in general. The variability of sales on sundays is quite high
while the median is not:

```{r}
ggplot(train[Sales != 0],
       aes(x = factor(DayOfWeek), y = Sales)) + 
    geom_jitter(alpha = 0.1) + 
    geom_boxplot(color = "yellow", outlier.colour = NA, fill = NA)

```

**The `store` file contains information about the stores that can be linked to 
`train` and `test` via the store ID.**

```{r}
summary(store)
table(store$StoreType)
table(store$Assortment)
# There is a connection between store type and type of assortment
table(data.frame(Assortment = store$Assortment, StoreType = store$StoreType))
hist(store$CompetitionDistance, 100)
# Convert the CompetitionOpenSince... variables to one Date variable
store$CompetitionOpenSince <- as.yearmon(paste(store$CompetitionOpenSinceYear, 
                                               store$CompetitionOpenSinceMonth, sep = "-"))
# One competitor opened 1900
hist(as.yearmon("2015-10") - store$CompetitionOpenSince, 100, 
     main = "Years since opening of nearest competition")
# Convert the Promo2Since... variables to one Date variable
# Assume that the promo starts on the first day of the week
store$Promo2Since <- as.POSIXct(paste(store$Promo2SinceYear, 
                                   store$Promo2SinceWeek, 1, sep = "-"),
                             format = "%Y-%U-%u")
hist(as.numeric(as.POSIXct("2015-10-01", format = "%Y-%m-%d") - store$Promo2Since), 
     100, main = "Days since start of promo2")
table(store$PromoInterval)
```

The stores with promos tend to make lower sales. This does not necessary mean
that the promos don't help or are counterproductive. They are possibly measures
that are taken mainly by stores with low sales in the first place:

```{r}
# Merge store and train 
train_store <- merge(train, store, by = "Store")
ggplot(train_store[Sales != 0], aes(x = factor(PromoInterval), y = Sales)) + 
    geom_jitter(alpha = 0.1) + 
    geom_boxplot(color = "yellow", outlier.colour = NA, fill = NA)
```

The different store types and assortment types imply different overall levels of sales and seem to
be exhibiting different trends:

```{r}
ggplot(train_store[Sales != 0], 
       aes(x = as.Date(Date), y = Sales, color = factor(StoreType))) + 
    geom_smooth(size = 2)
ggplot(train_store[Customers != 0], 
       aes(x = as.Date(Date), y = Customers, color = factor(StoreType))) + 
    geom_smooth(size = 2)
ggplot(train_store[Sales != 0], 
       aes(x = as.Date(Date), y = Sales, color = factor(Assortment))) + 
    geom_smooth(size = 2)
ggplot(train_store[Sales != 0], 
       aes(x = as.Date(Date), y = Customers, color = factor(Assortment))) + 
    geom_smooth(size = 2)
```

The effect of the distance to the next competitor is a little counterintuitive.
Lower distance to the next competitor implies (slightly, possibly not significantly)
higher sales. This may occur
(my assumption) because stores with a low distance to the next competitor are
located in inner cities or crowded regions with higher sales in general. Maybe
the effects of being in a good / bad region and having a competitor / not 
having a competitor cancel out:

```{r}
salesByDist <- aggregate(train_store[Sales != 0 & !is.na(CompetitionDistance)]$Sales, 
               by = list(train_store[Sales != 0 & !is.na(CompetitionDistance)]$CompetitionDistance), mean)
colnames(salesByDist) <- c("CompetitionDistance", "MeanSales")
ggplot(salesByDist, aes(x = log(CompetitionDistance), y = log(MeanSales))) + 
    geom_point() + geom_smooth()
```

A missing value for `CompetitionDistance` doesn't necessarily mean that there is
no competiton. Maybe that data was just not collected, yet. There is no obvious 
connection between sales and having `NA` as `CompetitionDistance`:

```{r}
ggplot(train_store[Sales != 0],
       aes(x = factor(!is.na(CompetitionOpenSinceYear)), y = Sales)) +
    geom_jitter(alpha = 0.1) +
    geom_boxplot(color = "yellow", outlier.colour = NA, fill = NA) +
    ggtitle("Any competition?")
```

So what happens if a competitor opens? In order to assess this effect we fetch
data from all stores that first have `NA` as `CompetitorDistance` and later some
value. **Only the month, not the date, of the opening of the competitor is known
so we need a rather large window to see the effect (100 days)**. 147 stores
had a competitor move into their area during the available time span. The 
competition leaves a 'dent' in the sales which looks a little different 
depending on the chosen `timespan` so I wouldn't like to argue about statistical
significance based on this plot alone. It's informative to look at anyway:

```{r}
# Sales before and after competition opens
train_store$DateYearmon <- as.yearmon(train_store$Date)
train_store <- train_store[order(Date)]
timespan <- 100 # Days to collect before and after Opening of competition
beforeAndAfterComp <- function(s) {
    x <- train_store[Store == s]
    daysWithComp <- x$CompetitionOpenSince >= x$DateYearmon
    if (any(!daysWithComp)) {
        compOpening <- head(which(!daysWithComp), 1) - 1
        if (compOpening > timespan & compOpening < (nrow(x) - timespan)) {
           x <- x[(compOpening - timespan):(compOpening + timespan), ] 
            x$Day <- 1:nrow(x)
            return(x)
        }
    }
}
temp <- lapply(unique(train_store[!is.na(CompetitionOpenSince)]$Store), beforeAndAfterComp)
temp <- do.call(rbind, temp)
# 147 stores first had no competition but at least 100 days before the end
# of the data set
length(unique(temp$Store))
ggplot(temp[Sales != 0], aes(x = Day, y = Sales)) + 
    geom_smooth() + 
    ggtitle(paste("Competition opening around day", timespan))
```


The seasonplot is adapted from [spsrini](https://www.kaggle.com/spsrini/rossmann-store-sales/seasonplot-month/files):
 
```{r}
temp <- as.data.frame(train)
temp$year <- as.factor(format(temp$Date, "%Y"))
temp$month <- as.factor(format(temp$Date, "%m"))
agg <- aggregate(Sales ~ ., data=temp[, c("Sales", "month" ,"year")], FUN=sum)
SalesTS <- ts(agg$Sales, start=2013, frequency=12)
col = rainbow(3)
seasonplot(SalesTS, col=col, year.labels.left = TRUE, pch=19, las=1)
```